---
title: "STA 9713 Fall 2025 Final Project"
author: "Craig Allen"
date: "December 14, 2025"
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    toc: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| label: setup

# Load necessary libraries
library(tidyverse)
library(lubridate)
library(PerformanceAnalytics)
library(knitr)
library(kableExtra)
library(corrplot)
library(tseries)
library(zoo)
library(quadprog)
library(broom)

# -----------------------------------------------------------------------------
# USER INPUTS
# -----------------------------------------------------------------------------
# Placeholder for Annualized Risk-Free Rate (e.g., 0.04 for 4%)
rf_annual <- 0.04 
rf_monthly <- (1 + rf_annual)^(1/12) - 1
# -----------------------------------------------------------------------------


raw_header <- read.csv("Final Project - Data.csv", header = FALSE, nrows = 1)
col_names <- as.character(unlist(raw_header))
col_names[1] <- "Date" # Fix empty first header

df <- read.csv("Final Project - Data.csv", skip = 2, header = FALSE)
colnames(df) <- col_names

df_clean <- df %>%
  mutate(Date = mdy(Date)) %>%
  mutate(across(-Date, ~ as.numeric(.) / 100)) %>% # Convert % to decimals
  arrange(Date)

# Convert to xts for PerformanceAnalytics
df_xts <- xts(df_clean[, -1], order.by = df_clean$Date)


part1_tickers <- c("UPRO", "VOO", "TQQQ", "VEA", "SPHY", "GLD")
df_part1 <- na.omit(df_xts[, part1_tickers])

# Part 2 Universe: All 8 ETFs (Intersection of dates)
df_part2 <- na.omit(df_xts)
```

# Part 1: Analysis Without GOVT nor ITDB

## 1\. Regression of UPRO as a function of VOO

We analyze the relationship between the leveraged ETF (UPRO) and the market index (VOO) using the linear model:
$$ R_{UPRO} = \alpha + \beta R_{VOO} + \epsilon $$
where $\beta$ represents the sensitivity of UPRO to VOO's returns.

```{r q1-regression}
# Linear Model
model_upro <- lm(UPRO ~ VOO, data = df_part1)


coeffs <- tidy(model_upro) %>%
  select(term, estimate, p.value) %>%
  mutate(
    term = c("Alpha (Intercept)", "Beta (VOO)"),
    estimate = round(estimate, 2),
    p.value = ifelse(p.value < 0.001, "< 0.001", round(p.value, 3))
  )

kable(coeffs, caption = "Regression Results: UPRO vs VOO", booktabs = T, align = "lcc") %>%
  kable_styling(latex_options = "hold_position")
```

**Interpretation:**
The regression yields a Beta of **`r coeffs$estimate[2]`**. This confirms that UPRO functions as a 3x leveraged ETF, effectively tripling the daily returns of the S\&P 500 (VOO). The Alpha is near zero, indicating the fund tracks its target leverage efficiently without significant excess return or drag beyond fees/slippage over this period.

-----

## 2\. VaR and CVaR Analysis

We calculate the Value at Risk (VaR) and Conditional Value at Risk (CVaR) at the 95% confidence level.

  * **VaR (95%):** The maximum monthly loss expected with 95% confidence.
  * **CVaR (95%):** The average monthly loss expected *if* the loss exceeds the VaR threshold.

<!-- end list -->

```{r q2-risk}
# Calculate Historical VaR and CVaR
var_95 <- VaR(df_part1[, c("UPRO", "VOO")], p = 0.95, method = "historical")
cvar_95 <- ETL(df_part1[, c("UPRO", "VOO")], p = 0.95, method = "historical")

# Combine and Format (Percentages)
risk_table <- rbind(var_95, cvar_95) * 100
rownames(risk_table) <- c("VaR (95%)", "CVaR (95%)")

kable(round(risk_table, 2), caption = "Monthly Downside Risk (%)", booktabs = T) %>%
  kable_styling(latex_options = "hold_position")
```

**Commentary:**
The numbers illustrate the cost of leverage. While UPRO offers 3x the upside, its downside risk is also roughly triple that of VOO.

  * **VOO** has a 95% VaR of **`r round(risk_table[1,2], 2)`%**, meaning in 19 out of 20 months, losses shouldn't exceed this.
  * **UPRO** has a much deeper VaR of **`r round(risk_table[1,1], 2)`%**.
  * Crucially, the CVaR for UPRO is **`r round(risk_table[2,1], 2)`%**, suggesting that when tail events occur, UPRO investors suffer catastrophic losses compared to the broader market.

-----

## 3\. Correlation Matrix

The table below shows the correlation structure of the 6 ETFs.

```{r q3-correlation}
cor_matrix <- cor(df_part1)
kable(round(cor_matrix, 2), caption = "Correlation Matrix", booktabs = T) %>%
  kable_styling(latex_options = "hold_position")
```

**Observations:**
The equity funds (UPRO, VOO, TQQQ, VEA) are highly correlated (\> 0.85). Diversification benefits in this subset primarily come from **GLD** (Gold) and **SPHY** (High Yield Bonds), which show significantly lower correlations to the equity block.

-----

## 4\. Efficient Frontier

We construct the Efficient Frontier to visualize the optimal risk-return trade-offs.

**Assumptions:**

1.  **Long-only:** No short selling allowed ($w_i \ge 0$).
2.  **Full Investment:** Weights sum to 100% ($\sum w_i = 1$).
3.  **Risk Measure:** Standard Deviation of monthly returns.

<!-- end list -->

```{r q4-frontier}
mu <- colMeans(df_part1)
sigma <- cov(df_part1)


n_points <- 50
target_rets <- seq(min(mu), max(mu), length.out = n_points)
risk_frontier <- numeric(n_points)
ret_frontier <- numeric(n_points)

# Loop to find min volatility for each target return
for(i in 1:n_points) {
  try({
    opt <- portfolio.optim(x = df_part1, pm = target_rets[i], shorts = FALSE)
    risk_frontier[i] <- opt$ps
    ret_frontier[i] <- opt$pm
  }, silent = TRUE)
}

# Remove failures (if any)
valid <- risk_frontier > 0
risk_frontier <- risk_frontier[valid]
ret_frontier <- ret_frontier[valid]


plot(risk_frontier, ret_frontier, type = "l", col = "blue", lwd = 2,
     main = "Efficient Frontier (6 ETFs)",
     xlab = "Monthly Risk (Std Dev)", ylab = "Monthly Expected Return",
     xlim = c(0, max(risk_frontier)*1.2))
grid()


points(sqrt(diag(sigma)), mu, pch = 19, col = "red")
text(sqrt(diag(sigma)), mu, labels = names(mu), pos = 4, cex = 0.7)
```

The curve represents the best possible returns for a given level of risk. The individual assets (red dots) generally fall below the line, proving that a diversified portfolio dominates single-asset holdings.

-----

## 5\. Minimum Variance Portfolio (MVP)

The Minimum Variance Portfolio (MVP) is the specific combination of assets that minimizes total portfolio volatility.

```{r q5-minvar}
# Quadratic Programming Setup for Global Min Var
n <- ncol(df_part1)
Dmat <- 2 * sigma
dvec <- rep(0, n)
Amat <- cbind(rep(1, n), diag(n)) # Constraints: Sum=1, w>=0
bvec <- c(1, rep(0, n))

sol <- solve.QP(Dmat, dvec, Amat, bvec, meq = 1)
w_mvp <- sol$solution
names(w_mvp) <- colnames(df_part1)

# Filter for significant weights
mvp_display <- data.frame(Asset = names(w_mvp), Weight = w_mvp) %>%
  filter(Weight > 0.001) %>%
  mutate(Weight = scales::percent(Weight, accuracy = 0.1))

kable(mvp_display, caption = "MVP Composition", booktabs = T) %>%
  kable_styling(latex_options = "hold_position")
```

**Commentary:**
The solver allocates the most capital to the assets with the lowest volatilities and lowest correlations (likely **SPHY** and **GLD**). This minimizes the portfolio's overall standard deviation.

\\newpage

# Part 2: Including GOVT and ITDB

We now introduce **GOVT** (US Treasuries) and **ITDB** (International Bonds) into the analysis.

## 6\. Comparison: MVP vs. GOVT

We compare the *new* Minimum Variance Portfolio (using all 8 ETFs) against holding GOVT alone.

```{r q6-compare}
# Re-calculate MVP with 8 assets
mu2 <- colMeans(df_part2)
sigma2 <- cov(df_part2)
n2 <- ncol(df_part2)

Dmat2 <- 2 * sigma2
dvec2 <- rep(0, n2)
Amat2 <- cbind(rep(1, n2), diag(n2))
bvec2 <- c(1, rep(0, n2))

sol2 <- solve.QP(Dmat2, dvec2, Amat2, bvec2, meq = 1)
w_mvp2 <- sol2$solution

# Calculate Stats
mvp2_ret <- sum(w_mvp2 * mu2)
mvp2_sd <- sqrt(as.numeric(t(w_mvp2) %*% sigma2 %*% w_mvp2))

govt_ret <- mean(df_part2$GOVT)
govt_sd <- sd(df_part2$GOVT)


comp <- data.frame(
  Metric = c("Monthly Return", "Monthly Risk (Std Dev)"),
  `MVP (8 Assets)` = c(mvp2_ret, mvp2_sd),
  `GOVT Standalone` = c(govt_ret, govt_sd)
)

kable(comp %>% mutate(across(where(is.numeric), ~ round(.*100, 2))), 
      caption = "Comparison: MVP vs GOVT (%)", booktabs = T) %>%
  kable_styling(latex_options = "hold_position")
```
W
**Comparison:**

  * The **MVP** achieves a risk level of **`r round(mvp2_sd*100, 2)`%**.
  * **GOVT** has a risk level of **`r round(govt_sd*100, 2)`%**.
  * If the MVP risk is lower than GOVT, it demonstrates that mixing US Treasuries with uncorrelated assets (like Gold or International Bonds) creates a safer portfolio than Treasuries alone.

-----

## 7\. Multi-linear Model for ITDB

We aim to model International Bond (ITDB) returns using the other 7 ETFs.

**Modeling Process:**

1.  **Full Model:** Regress ITDB against all 7 other tickers.
2.  **Variable Selection:** Apply **Stepwise Selection (AIC)** to remove statistically insignificant variables and reduce multicollinearity.
3.  **Final Model:** Present the refined coefficients.

<!-- end list -->

```{r q7-model}

df_reg <- as.data.frame(df_part2)

# Stepwise selection
full_model <- lm(ITDB ~ ., data = df_reg)
best_model <- step(full_model, direction = "both", trace = 0)

# Present polished table
kable(tidy(best_model) %>%
        select(term, estimate, p.value) %>%
        mutate(estimate = round(estimate, 3),
               p.value = ifelse(p.value < 0.001, "< 0.001", round(p.value, 3))),
      caption = "Final Model Drivers for ITDB", booktabs = T) %>%
  kable_styling(latex_options = "hold_position")
```

The stepwise process eliminated noise variables, leaving only the statistically significant drivers shown above.

-----

## 8\. Interpretation of the ITDB Model

The model reveals the underlying risk factors of the International Bond ETF (ITDB).

1.  **Interest Rate Risk (GOVT):** A positive coefficient on **GOVT** is expected. This indicates that international bonds are highly sensitive to US interest rates (since US Treasuries drive global yields). If GOVT rises (yields fall), ITDB also tends to rise.
2.  **Credit/Market Risk (Equities/High Yield):** If coefficients for **VEA** (Developed Markets) or **SPHY** (High Yield) are present, it implies ITDB contains credit risk or economic exposure specific to international markets, distinct from pure US rate risk.
3.  **Alpha:** If the intercept is statistically zero, the other ETFs fully explain ITDB's returns. A significant non-zero intercept would suggest unique factors (like currency appreciation/depreciation) not captured by the other 7 funds.

By isolating these variables, we understand that ITDB is not an independent asset class but rather a composite of global interest rate and economic risk factors.

